---
title: "Exploring the tidyHeatmap R package"
author: "Thomas Sandmann"
format: html
editor_options: 
  chunk_output_type: console
---

## Installing the tidyHeatmap R package

The `tidyHeatmap` R package is available from
[CRAN](https://cran.r-project.org/package=tidyHeatmap) [^1].

[^1]: Because `tidyHeatmap` relies heavily on other packages from the
[tidyverse](https://www.tidyverse.org/),
e.g. the 
[dplyr](https://dplyr.tidyverse.org//)
and 
[tidyr](https://tidyr.tidyverse.org//)
R packages, it has a relatively deep dependency tree. But that's 
par for the course if you want to adopt the 
[tidy tools API](https://tidyverse.tidyverse.org/articles/manifesto.html) 
by using Posit's toolchain.

The latest release available from CRAN can be installed via: 

```{r}
#| eval: false
install.packages("tidyHeatmap")
```

and the latest development version (used in this tutorial) is available
from github:

```{r}
#| eval: false
remotes::install_github("stemangiola/tidyHeatmap")
```

🚨 At the time of writing, the latest version available from CRAN (version 
1.8.1) is outdated; please install & use the latest development version from
github (currently version 1.10.1) to follow along with this tutorial.

## Targeted lipidomics data published by Xia et al

In this tutorial we will used targeted mass spectrometry data 
published by 
[Xia et al, Molecular neurodegeneration, 2022](https://molecularneurodegeneration.biomedcentral.com/articles/10.1186/s13024-022-00547-7).
Supplementary table 5 of this publication contains the normalized abundances for
151 analytes measured across 18 samples. We start by downloading the 
supplementary excel file and wrangle sample annotations, feature annotations
and the measurements into a tidy tibble with the `get_published_data()` 
functxion defined in the callout box below.

::: {.callout-tip  collapse="true"}

## Retrieving & tidying published LC/MS data

```{r}
#| message: false

library(checkmate)
library(dplyr)
library(readxl)
library(tidyr)

#' @noRd
#' @param path Scalar character, the path to the supplementary excel file
#' @importFrom checkmate assert_file
#' @importFrom readxl read_excel
.parse_sample_annotations <- function(path) {
  checkmate::assert_file(path)
  # parse sample- and feature-annotations, and normalized abundances
  df <- readxl::read_excel(path, sheet = "sample_annotations")
  df[, c("sample_id", "cell_number", "genotype", "sex", "batch")]
}

#' @noRd
#' @param path Scalar character, the path to the supplementary excel file
#' @importFrom checkmate assert_file
#' @importFrom readxl read_excel
.parse_feature_annotations <- function(path) {
  checkmate::assert_file(path)
  df <- readxl::read_excel(path, sheet = "feature_annotations")
  df <- df[, grep("QTRAP|XEVO|m.z", colnames(df), invert = TRUE)]
  df <- df[!duplicated(df$component_name), ]
  df$feature_id <- NULL
  df <- df[!df$is_internal_standard,]
  df[, setdiff(
    colnames(df), c("is_internal_standard", "panel")),
    drop = FALSE]
}

#' @noRd
#' @param path Scalar character, the path to the supplementary excel file
#' @importFrom checkmate assert_file
#' @importFrom tidyr pivot_longer
#' @importFrom readxl read_excel
.parse_abundances <- function(path) {
  checkmate::assert_file(path)
  df <- readxl::read_excel(path, sheet = "peak_area_ratio_to_is")
  df <- data.frame(
    component_name = df$component_name,
    df[, grep("^LA", colnames(df), value = TRUE)]
  )
  tidyr::pivot_longer(df, cols = setdiff(colnames(df), "component_name"),
                      names_to = "sample_id", values_to = "abundance")
}

#' Retrieve targeted LC/MS data published by Xia et al, 2022
#' @importFrom dplyr left_join join_by
#' @export
#' @return A tidy tibble with 2,718 rows and 9 columns
#' @source [Xia et al, Molecular neurodegeneration, 2022](https://molecularneurodegeneration.biomedcentral.com/articles/10.1186/s13024-022-00547-7)
#' @examples
#' get_published_data()
get_published_data <- function() {
  url <- paste0(
    "https://static-content.springer.com/esm/art%3A10.1186%2",
    "Fs13024-022-00547",
    "-7/MediaObjects/13024_2022_547_MOESM13_ESM.xlsx")
  
  # retrieve the first supplementary data file
  tmp_file <- tempfile(fileext = ".xlsx")
  download.file(url, tmp_file, quiet = TRUE)
  # parse sample- and feature-annotations & normalized abundances
  sample_anno <- .parse_sample_annotations(tmp_file)
  feature_anno <- .parse_feature_annotations(tmp_file)
  abundances <- .parse_abundances(tmp_file)
  # join all tables into a single tidy tibble
  dplyr::left_join(
    feature_anno, abundances, by = dplyr::join_by("component_name")) |>
    dplyr::left_join(sample_anno, by = dplyr::join_by("sample_id")) |>
    dplyr::mutate(
      sample_id = factor(sample_id, levels = paste0("LA", 1:18, "C")),
      sex = factor(sex),
      ionization = factor(ionization),
      batch = factor(batch),
      genotype = factor(genotype, levels = c("WT", "Het", "Hom"))
    )
}
```
:::

```{r}
df <- get_published_data()  # a tidy tibble
```

Now we have the data in a tidy tibble with `r nrow(df)` rows and `r ncol(df)`
columns.

To control the order of categorical variables (e.g. `genotype` or or `batch`),
they are encoded as factors.[^2] 

[^2]: For example, the three possible values of `genotype` are ordered according
to the number of copies of the APP-SAA  transgene: `WT` (zero copies), 
`Het` (1 copy) and `Hom` (2 copies).

```{r}
summary(df)
```

## A first heatmap

The `tidyHeatmap` R package provides wrapper functions that make the
functionality of the
[ComplexHeatmap Bioconductor package](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html)
available within a tidy workflow. [^3]

[^3]: The documentation of the `ComplexHeatmap` package is amazing: the
[Complete Reference](https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html) contains to 15 chapters.

```{r}
#| message: false

library(circlize)  # to create continuous color gradients
library(grid)  # to define witdth / height of annotations using grid::unit()
library(ragg)  # to generate high quality graphics
library(RColorBrewer)  # to access predefined color palettes
library(tidyHeatmap)
```

Creating a first heatmap is as easy as passing our tidy tibble into the 
`tidyHeatmap::heatmap()` function, and specifying which columns contain
row identifiers, column identifiers and values:

```{r}
heatmap(.data = df, .row = component_name, .column = sample_id, 
        .value = abundance, scale = "row")
```

::: {.callout-tip  collapse="false"}

The normalized abundances are hard to interpret (because analytes were
normalized to different internal standards and don't correspond to absolute
concentrations). Luckily, we are mainly interested in more interested in the
relative changes across samples (and groups of samples). By setting 
`scale = "row"`, the values for each analyte are converted to z-scores - and
the color scale indicates variation around the mean of each row.

:::

We can pass additional arguments through to the `ComplexHeatmap::Heatmap()`
function, e.g. specify another method to cluster rows via the 
`clustering_method_rows` argument, by suppressing column clustering entirely
via `cluster_columns=FALSE`, or by defining the color palette to use [^4].

[^4]: See 
[Heatmap's help page](https://www.rdocumentation.org/packages/ComplexHeatmap/versions/1.10.2/topics/Heatmap) 
for a full list of arguments.

```{r}
colors <- circlize::colorRamp2(
            seq(-2, 2, length.out = 11), 
            RColorBrewer::brewer.pal(11, "RdBu")
        )
heatmap(.data = df,
        .row = component_name, 
        .column = sample_id, 
        .value = abundance, 
        column_title = "Xia et al: targeted LC/MS data",
        row_title = "Analyte",
        scale = "row",
        cluster_columns = FALSE,
        clustering_method_rows = "ward.D",
        clustering_method_columns = "ward.D",
        palette_value = colors
)
```
We can also introduce visual splits between the columns, e.g. separating
the three genotypes from each other, and assign a different color to each one:

```{r}
df |>
  group_by(genotype) |>
  heatmap(.row = component_name, 
          .column = sample_id, 
          .value = abundance, 
          column_title = "Xia et al: targeted LC/MS data",
          row_title = "Analyte",
          scale = "row",
          cluster_columns = FALSE,
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D",
          palette_value = colors,
          palette_grouping = list(
            c("darkgrey", "orange", "firebrick")
        )
  )
```
## Splitting features

The dendrogram (on the left) indicates that there might be two major groups
of features in this dataset, e.g. there are two main branches. 

### Cutting the hierarchical clustering result

We can reinforce this result by cutting the dendrogram after two branches have
arisen.

```{r}
df |>
  group_by(genotype) |>
  heatmap(.row = component_name, 
          .column = sample_id, 
          .value = abundance, 
          column_title = "Xia et al: targeted LC/MS data",
          row_title = "Analyte",
          scale = "row",
          cluster_columns = FALSE,
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D",
          palette_value = colors,
          palette_grouping = list(
            c("darkgrey", "orange", "firebrick")
        )
  ) |>
    split_rows(2)
```

::: {.callout-tip  collapse="true"}

### K-means clustering

Alternatively, we can can perform
[k-means clustering](https://en.wikipedia.org/wiki/K-means_clustering)
on the rows (or columns) by specifying the number of clusters (k) to isolate
via the `row_km` argument. Our current dataset contains missing values 
(`NA`), which is incompatible with the k-means clustering algorithm. To use this
method we first need to decide how to handle missing values, e.g. drop analytes
or impute reasonable values.

:::

## Adding annotations

The `annotation_tile()` function[^5] allows us to add additional annotations to
our heatmap, e.g. to identify 

- columns from the `batches` or correspoding to animals with male or female 
  `sex` in an annotation bar on top of the heatmap, or
- rows with values from different `ionization` conditions in an annotation
  strip on the left of the heatmap.

[^5]: Prior to `tidyHeatmap` version 1.9.0, the `add_tile()` function was used
instead. It is now deprecated in favor of `annotation_tile()`. (At the time of
writing, the latest version available from CRAN is version 1.8.1; please
install & use the latest development version from github to follow along this
tutorial.)


```{r}
df |>
  group_by(genotype) |>
  heatmap(.row = component_name, 
          .column = sample_id, 
          .value = abundance, 
          column_title = "Xia et al: targeted LC/MS data",
          row_title = "Analyte",
          scale = "row",
          cluster_columns = FALSE,
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D",
          palette_value = colors,
          palette_grouping = list(
            c("darkgrey", "orange", "firebrick")
        )
  ) |>
    annotation_tile(
        batch, 
        palette = RColorBrewer::brewer.pal(3, "Dark2")
    ) |>
    annotation_tile(sex, 
             palette = c("skyblue", "pink")) |>
    annotation_tile(ionization, size = grid::unit(5, "points"))
```

::: {.callout-tip  collapse="true"}

### Row or column annotations?

The `annotation_tile()` function does not require us to specify whether we
are adding row or column annotations. Instead, it infers it automatically:

- Does each value used to define a row of the heatmap (e.g. `component_name`)
  map to a single value in the selected column of the data.frame?
  - If yes, create a row annotation strip.
- Does each value used to define a column of the heatmap (e.g. `sample_id`)
  map to a single value in the selected column of the data.frame?
  - If yes, create a column annotation.
- If neither of the above are true, an error is raised.

:::

## Adding quantitative annotations

Some annotations are quantitative, and may be better communicated in a plot.
Let's display the number of cells that were analyzed in each sample on top of
our heatmap with the `add_point()` function:

```{r}
df |>
  group_by(genotype) |>
  heatmap(.row = component_name, 
          .column = sample_id, 
          .value = abundance, 
          column_title = "Xia et al: targeted LC/MS data",
          row_title = "Analyte",
          scale = "row",
          cluster_columns = FALSE,
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D",
          palette_value = colors,
          palette_grouping = list(
            c("darkgrey", "orange", "firebrick")
          )
  ) |>
  annotation_tile(
    batch, 
    palette = RColorBrewer::brewer.pal(3, "Dark2")
  ) |>
  annotation_tile(
    sex, 
    palette = c("skyblue", "pink")
  ) |>
  annotation_tile(ionization, size = grid::unit(5, "points"))
```

## Overlaying symbols

To highlight specific cells in our heatmap, the `layer_point()`, 
`layer_arrow_up()`, `layer_arrow_down()`, `layer_diamond()`, `layer_square()`,
`layer_star()`, `layer_text()` or `layer_asterisk()`
functions can be used, together with a filter condition [^5].

[^6]: At the time of writing, the latest version of the `tidyHeatmap` package
available from CRAN is version 1.8.1. The development version used here, 
available [from github](https://github.com/stemangiola/tidyHeatmap),
contains additional functionality, including e.g. the `layer_star()` and - most
importantly - `layer_asterisk()`.

```{r}
df |>
  group_by(genotype) |>
  heatmap(.row = component_name, 
          .column = sample_id, 
          .value = abundance, 
          column_title = "Xia et al: targeted LC/MS data",
          row_title = "Analyte",
          scale = "row",
          cluster_columns = FALSE,
          clustering_method_rows = "ward.D",
          clustering_method_columns = "ward.D",
          palette_value = colors,
          palette_grouping = list(
            c("darkgrey", "orange", "firebrick")
        )
  ) |>
  annotation_tile(
    batch, 
    palette = RColorBrewer::brewer.pal(3, "Dark2")
  ) |>
  annotation_tile(sex, 
                  palette = c("skyblue", "pink"))  |>
  annotation_tile(ionization, size = grid::unit(5, "points")) |>
  layer_asterisk(
    component_name == "PC(40:6)" & sample_id %in% c("LA2C", "LA10C")
  )
```

## Reproducibility

<details>
<summary>
Session information
</summary>

```{r}
sessionInfo()
```
</details>
